---
title: "Coursera Reproducible Research Assignment"
output: html_document
---
###This report is the result of a Cousera Reproducible Research Assignment to analyse a person's movement based on data from an activity monitoring device. The dataset used in the report was obtained from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip (accessed 13-Feb-2016).The objective of the report is to demonstrate to usefulness of R Markdown in Reproducible research.


### 1. Loading and processing of the data
```{r eval=TRUE }
library(dplyr)
library(ggplot2)
```

```{r echo = TRUE}
# read unzipped csv file and convert date field to date format
activitydata <- read.csv("./activity.csv")
activitydata$date <- as.Date(activitydata$date)
```

### 2. Calculating the total number of steps taken per day.

```{r echo = TRUE}
totals_steps_day <- filter(activitydata, !is.na(steps)) %>% 
		group_by(date) %>% 
		summarise(total_steps = sum(steps))
totals_steps_day
```

### 3. Histogram of the total number of steps taken each day (excluding missing values)
```{r echo = TRUE}
hist(totals_steps_day$total_steps, 
	main="Histogram of Total Steps taken daily - Missing values excluded",
	xlab="Total Steps",
	ylim = c(0,40),
	border="blue",
	col="green")
```	


	
### 4. Calculate and report the mean and median of the total number of steps taken per day

```{r echo = TRUE}
mean_median_steps_day <- filter(activitydata, !is.na(steps)) %>% 
		group_by(date) %>% 
		summarise(mean_steps = mean(steps), median_steps = median(steps))
mean_median_steps_day
```


### 5. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) the average number of steps taken, averaged across all days (y-axis)

```{r echo = TRUE}
avg_interval_steps <- filter(activitydata, !is.na(steps)) %>% 
		group_by(interval) %>% 
		summarise(avg_steps = mean(steps))

plot(avg_interval_steps$interval,avg_interval_steps$avg_steps,
	type="l",
	main="5-minute interval vs Avg steps taken",
	xlab="5-Minute Interval",
	ylab="Avg Steps Taken")
```



### 6. Determining the 5-minute interval that, on average, contains the maximum number of steps

```{r echo = TRUE}
max_interval_steps <- filter(avg_interval_steps, avg_steps==max(avg_interval_steps$avg_steps)) %>% 
					select(interval)
max_interval_steps
```

The 5-minute interval with the maximum average number of steps is:
```{r }
max_interval_steps$interval
```



### 7. The strategy used to impute the missing values is as follows:  Using a copy of the activitydata data a lookup field is created by combining the day of the week and 5-minute interval number eg.Monday-5, Monday-10, Sunday-100 etc. Using this field another data frame was created with the lookup field and the average number of steps for each day of week-5minute interval combination. The two data frames are then merged by the lookup field and the NA values replaced with this average value. By using this strategy a more accurate estimate of the missing values can be obtained.

#### The total number of missing values in the dataset is:
```{r echo = TRUE}
cnt_na <- NROW(activitydata[is.na(activitydata$steps),])
cnt_na
```

#### Executing the strategy
```{r echo = TRUE}
#Copy data set and add a lookup field
activitydata2 <- activitydata
activitydata2$lookup <- paste(weekdays(activitydata2$date),"-",activitydata2$interval)

#Create data frame with average steps for lookup field
avg_interval_steps_day <- filter(activitydata2, !is.na(steps)) %>% 
		group_by(lookup) %>% 
		summarise(avg_steps = round(mean(steps),0)) 

#Merge data sets by lookup field
merged <- merge(activitydata2,avg_interval_steps_day, by.x = "lookup", by.y = "lookup",sort = FALSE)

#Replace missing values with average from lookup field
merged$steps_new <- ifelse(is.na(merged$steps),merged$avg_steps,merged$steps)
```


### 8. Histogram of the total number of steps taken each day after missing values are imputed 

```{r echo = TRUE}
##mean and median total number of steps taken per day.

totals_steps_day2 <- filter(merged, !is.na(steps_new)) %>% 
		group_by(date) %>% 
		summarise(total_steps = sum(steps_new))

# Histogram of the total number of steps taken each day
hist(totals_steps_day2$total_steps, 
	main="Histogram of Total Steps taken daily - Missing values Imputed",
	xlab="Total Steps",
	ylim = c(0,40),
	border="blue",
	col="green")
```


#### Calculating mean and median of total number of steps taken daily
```{r echo = TRUE}
    mean_median_steps_day2 <- filter(merged, !is.na(steps_new)) %>% 
		group_by(date) %>% 
		summarise(mean_steps = mean(steps_new), median_steps = median(steps_new))
    mean_median_steps_day2
```




### 9. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r echo = TRUE}
#Insert column to hold day type
merged$day_type <- ifelse(weekdays(merged$date) %in% c("Saturday","Sunday"),"Weekend","Weekday")

#summarise data by day type
day_type_summary <- merged %>% group_by(day_type,interval) %>% 
		summarise(avg_steps = mean(steps_new))

#plot chart
qplot(interval,avg_steps, 
	data = day_type_summary, 
	facets = day_type~. , 
	geom = c("line"),
	xlab = "5-minute Interval",
	ylab = "Average Number of Steps",
	main = "Comparison of Weekday and Weekend Activity patterns")
```


